Poker Chips, Earthquakes, and Computation

### 2014-10-13 finaldraft.tex
merging

In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the problem. In particular, we formalize the effects of earthquakes, as well as what it means for a board of chips to be finite or infinite. 
In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to positive integers. 
We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations. We also show that the number of possible configurations a finite board can be transformed into by earthquake is bounded. These two results together will show that all finite boards reach a steady state. 
In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.

For the remainder of the paper, we consider other approaches and extensions of the problem. 
%In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes. In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
%We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
In Section~\ref{Board Variants: d-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
\section{Definitions and Notation}
\label{Definitions and Notation}

@@

\begin{figure}
\label{cartesiangrid}
\begin{center}
\includegraphics[scale=0.8]{cartesiangrid.png}
\end{center}
\caption{A finite component of the grid with coordinates}
\end{figure}
We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$. We will define the board as a function from the Cartesian grid to the nonnegative integers.

@@

\begin{figure}
\centering
\subfigure[Board before earthquake.]{\includegraphics[width=0.25\linewidth]{action1.pdf}\label{fig:boardbeforeearthquake}}
\qquad\qquad
\subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
\caption{Earthquake applied to board. Green signifies nodes with more than 4 chips}
\label{fig:earthquakeexample}
\end{figure}

@@

Now that we have formalized the earthquake operator, we make precise the definition of stability.
\begin{definition}
A board is stable if all nodes contain values strictly less than 4. 
\end{definition}
This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be changed by earthquake operations.

We also formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.

@@

Finally, we use the above definitions to define global earthquakes, which will be defined only for finite boards. 
\begin{definition}
Given a finite board $b$, a global earthquake is given by applying one earthquake to every square in $b$ that initially has 4 or more chips.
\end{definition}

We will see in Section~\ref{Order of Redistribution} that the order in which earthquakes are applied does not matter, and thus global earthquakes are well-defined without the need for an ordering on the individual earthquakes. In order to avoid the case where there may be infinitely many squares with 4 or more chips, we only define global earthquakes for finite boards.

@@

In this section, we show that any finite board will become stable after a finite number of earthquakes. We first introduce a spread function that maps boards to the positive integers. We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration. In other words, we can never return to a given board arrangement.

We then give a bound on the maximum number of board configurations a finite board can be transformed into by earthquakes. We will argue that since a finite number of boards are reachable and board states are never revisited, eventually this process must terminate. Thus, we eventually reach a stable board where no earthquakes will change the board.

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\begin{equation*}
\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
\end{equation*}

Essentially the function sums up the number of chips on a given node weighted by the taxicab distance to the origin squared. The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are. Intuitively, earthquakes takes chips that are condensed in a single pile and spreads them out over a greater area. We capture this in the following lemma:

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb \neq b$. Then
\begin{equation*}
\Psi(E_sb) > \Psi(b) 
\end{equation*}
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

Let the taxicab distance to $s$ be $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.

Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$. In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$. If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$. We only need a lower bound for the increase in the spread function, so we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.

Suppose that the two neighboring nodes at distance $d_s+1$ have chip stacks of $n_1$ and $n_2$ originally. Then their contribution to the spread function goes from $n_1(d_s+1)^2$ and $n_2(d_s+1)^2$ to $(n_1+1)(d_s+1)^2$ and $(n_2+1)(d_s+1)^2$. If the other two nodes at distance $d_s-1$ have chip stacks of $n_3$ and $n_4$, their contribution to the spread function goes from $n_3(d_s-1)^2$ and $n_4(d_s-1)^2$ to $(n_3+1)(d_s-1)^2$ and $(n_4+1)(d_s-1)^2$.

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\begin{equation*}
 \Psi(E_sb) - \Psi(b) = 4
 \end{equation*}
which completes the proof.
\end{proof}

\begin{corollary}
\label{norepeats}
A finite board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}

Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that earthquakes can transform a given finite board into.
\begin{lemma}
\label{finiteextension}
Suppose there are a total of $C$ chips on the board, and $b$ is nonzero in the region $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$ (equivalently, all chips on the board are in this region). Then after any sequence of earthquakes, the resultant board is nonzero only in the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
\end{lemma}

\begin{proof}
This follows from the observation that if a row / column of the board contains at least one chip, it will always contain at least one chip. Note that this follows from the fact that an earthquake operation can never cause a board to break this property; if a row / column is affected by an earthquake operation, at least 2 chips stay inside that row / column.

Now consider the maximum value $x$ coordinate of the board that chips can ever ``reach". An earthquake must first move a chip to the column given by $x$ coordinate $i_{max} +1$, and a chip will always stay there. The same argument follows for $i_{max} + 2$, and so on, but we can only do this up until $i_{max} +C$. The same argument works in any direction, so we can never leave the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
\end{proof}

We are now ready to show that all finite boards will reach a stable state.

\begin{theorem} If $b$ is a finite board that is repeatedly hit by earthquakes, it must eventually reach a stable state. 
\end{theorem}
\label{finitestability}
\begin{proof}
Lemma~\ref{finiteextension} implies that there a finite board with $C$ total chips can only ever occupy a finite number $D$ of squares. This implies a finite upper bound of $\binom{D+C-1}{D-1}$ possible boards (via an elementary combinatorics argument of how to distribute $C$ identical objects among $D$ distinguishable slots).

In addition, Corollary~\ref{norepeats} tells us that we never visit a board state more than once. Thus, since we have a finite number of reachable board states, and a nontrivial earthquake operation always outputs a previously unvisited board, a sequence of earthquake operations must eventually end. At this end state, the board cannot allow further nontrivial earthquakes, and is stable by definition.
\end{proof}

@@

\begin{proof}
We will proceed by contradiction. Assume otherwise, so there exists some sequence $\{s_1, s_2, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $E_s b$ the resultant board is $b$.
However, we can bound the size of the locations affected by $\{s, s_1, s_2, \ldots, s_k \}$, call it $R$.  This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$. But this means that earthquakes transformed a finite board back into a previous state which contradicts Lemma ~\ref{finiteextension}. Therefore infinite boards cannot be transformed back into a previous state.
\end{proof}

@@

In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur. We show this by first describing conditions under which we can rearrange a sequence earthquakes without changing the resultant board. We then define a process that rearranges a sequence of earthquakes without changing the final board state and for a given board will always produce the same sequence.\\

@@

\begin{definition}
We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
\end{definition}

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$ if $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2 E_1 b = E_1 E_2 b
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma ~\ref{earthquakeredefine} we have 
\begin{align*}
E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1 E_2 b(x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

@@

Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ applied to $b_{m-1}$.

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in lemma ~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma ~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction we used in our first step of this proof we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order. 
The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, based on the initial board and then move these earthquakes to the front of the sequence without changing the final board. This technique will specify an order in which the earthquakes will occur, depending only on the original board which completes the proof.

\begin{theorem}
\label{thm:order}
Let $b$ be a board that becomes a stable board $b$ after $N$ earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$. Furthermore the number of times that a node $s$ is targeted is invariant of the order in which the earthquakes occur. 
\end{theorem}
\begin{proof}
Given the original configuration of the board, let $A_i$ be the number of chips on node $i$. We know that node $i$ will be hit by at least 
\begin{equation*}
\lfloor \frac{A_i}{4} \rfloor
\end{equation*}
earthquakes in order to stabilize. Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$. We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a node that starts with $A_i$ chips on it. Thus using Lemma ~\ref{shiftlemma}, we can move these earthquakes to the front of the sequence without changing the final positioning of the board. We do this for all nodes $i$.

The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the nodes.

Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the nodes. We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes. We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma ~\ref{shiftlemma}).

Thus we have transformed our starting sequence into a sequence that is completely determined by the state of the initial board, noting that the final state of the board is still the same.  We can do this for any other sequence that stabilizes the board and get the same sequence. This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them). 
\end{proof}

Note that the results of this section imply that global earthquakes are a well-defined operation on finite boards. Recall that a global earthquake is an operation defined on finite boards that hits every square of a finite board with 4 or more chips with an earthquake. By Theorem~\ref{thm:order}, the order of standard earthquakes does not matter, and thus global earthquakes are unambiguously defined. This is important as we will refer to global earthquakes in Section~\ref{Computing with Chips}.

@@

%\section{Time until Stability}
%\label{Time until Stability}
%Another question of interest is how many earthquakes must be applied to a finite board in order for it to stabilize. In order to simplify our simulations and analysis, we restricted our attention to global earthquakes, which hit entire boards and lead to a high degree of symmetry. We decided that this restriction was reasonable, as global earthquakes 
%We will confine our analysis to the case of the board with only one stack containing $n$ chips. We are able to demonstrate upper bounds on the number of earthquakes it takes to stabilize the board and we conjecture on the existence of tighter ones.  
%Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
%\begin{definition}
%We define $T(n)$ to be the number of global earthquakes that must be applied to the board for it to stabilize.
%We can get an upper board for $T(n)$ by looking at the maximum value of our spread function.
%Recall that by Lemma ~\ref{finiteextension} after any number of earthquakes the resultant board will be nonzero only on nodes in the range $[-n, n] \times [-n, n]$. 
%So we know that when the board is stable at some $b'$, 

>>>>

In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the problem. In particular, we formalize the effects of earthquakes, as well as what it means for a board of chips to be finite or infinite. 
In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to $\mathbb{R}$. 
We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state. Then we show that this means that no boards can return to their same state. 
From this, we conclude that such boards never cycle back to previous configurations. 
In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.

For the remainder of the paper, we consider other approaches and extensions of the problem. 
In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes. In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
In Section~\ref{Board Variants: d-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits. We will also give impossibility results that show the limits of our model of computation.
\section{Definitions and Notation}
\label{Definitions and Notation}

@@

\begin{figure}
\label{cartesiangrid}
\begin{center}
\includegraphics[scale=0.8]{cartesiangrid.png}
\end{center}
\caption{A finite component of the grid with coordinates}
\end{figure}

We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$. We will define the board as a function from the Cartesian grid to the nonnegative integers.

@@

\begin{figure}
\centering
\subfigure[Board before earthquake.]{\includegraphics[width=0.25\linewidth]{action1.pdf}\label{fig:boardbeforeearthquake}}
\qquad\qquad
\subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
\caption{Earthquake applied to board. Green signifies active nodes}
\label{fig:earthquakeexample}
\end{figure}

@@

Now that we have formalized the earthquake operator, we make precise the definition of stability.
\begin{definition}
A board is stable if all nodes contain values strictly less than 4. 
\end{definition}
This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be affected by earthquake operations.

Finally, we formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.

@@

@@

In this section, we show that any finite board will eventually become stable after a finite number of earthquakes. We first introduce a spread function that maps boards to positive numbers. We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration. In other words, we can never return to a given board arrangement.

We then give a bound on the maximum number of board configurations a finite board can ever reach. We will argue that since a finite number of boards are reachable and board states are never revisited, that eventually this process must terminate. Thus, we eventually reach a stable board where no earthquakes are allowable.

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]

The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are. Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area. This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.

\begin{theorem}
\label{finitestability}
If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
\end{theorem}

In order to show finite boards never revisit a prior arrangement, we define a spread function that maps boards to nonnegative integers. The spread function should give some measure of how ``spread out" the chips on a board are. Earthquakes seem to ``spread out" the chips on a board and so we would like earthquakes to increase the value of the function.

\begin{lemma}
Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
\end{lemma}

\begin{proof}
For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
\begin{equation*}
S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
\end{equation*}
We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared. This is well-defined since the total number of chips on the board is finite. We claim that 
\begin{equation*}
E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
\end{equation*}
That is if an earthquake changes the board, $S$ increases. This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.

We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
\begin{align*}
S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
\end{align*}
For at most two of the neighbors of $(s_1, s_2)$ 
\begin{equation*}
(|i|+|j|)= (|s_1|+|s_2|-1)^2,
\end{equation*}
and for at least two of the neighbors 
\begin{equation*}
(|i|+|j|)=(|s_1|+|s_2|+1)^2,
\end{equation*}
 and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
\begin{align*}
&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
\end{align*}
\end{proof}

Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that earthquakes can transform a given finite board into.
\begin{lemma}
\label{finiteextension}
Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then after applying any number of earthquakes the resultant board is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
\end{lemma}

\begin{proof}
Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E_{s}b(i,j) \geq 1$. This is clear since an earthquake can bring things into column $i$ or if it moves some chips out, then there are two chips that remain in the column.

The same is true for each row. This means that if a chip  is ever in a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
\end{proof}

We are now ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.  Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$. If we set 
\begin{equation*}
D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
\end{equation*}
then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,. We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
\end{proof}

@@

\begin{proof}
We will proceed by contradiction. Assume otherwise, so there exists some sequence $\{s_1, s_2, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $E_s b$ the resultant board is $b$.
However, we can bound the size of the locations affected by $\{s, s_1, s_2, \ldots, s_k \}$, call it $R$.  This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$. But this means that earthquakes transformed a finite board back into a previous state which is a contradiction. Therefore infinite boards cannot be transformed back into a previous state.
\end{proof}

@@

In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur. We do so by first describing conditions under which the position of an Earthquake in a sequence can change without effecting the final state. We then define a process that rearranges a sequence of earthquakes without changing the final board state and produces the same sequence for any board.\\

@@

\begin{definition}
We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
\begin{equation*}
E^{m}[b, \{s_k\}](s_{m+1}  \geq 4
\end{equation*}
for all $1\geq m+1 \leq n$.
\end{definition}

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$ if $E$ is a valid earthquake, then 
\begin{equation*}
E[b, s](x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2[E_1[b, s_1] s_2] = E_1[LE_2[b, s_2], s_1] 
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[b, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1[b, s_1](s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma ~\ref{earthquakeredefine} we have 
\begin{align*}
E_2[E_1[b, s_1], s_2](x) = E_1[b, s_1])(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2[b, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1[E_2 [b, s_2], s_1](x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

@@

Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, LE_j$ applied to $b_{m-1}$.

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in lemma ~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma ~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction as above  we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order. 
The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board. This technique will then specify an order in which the earthquakes will occur, which completes the proof.

\begin{theorem}
\label{thm:order}
Let $b$ be a board that becomes a stable board $b$ after $N$ earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$. Furthermore the number of times that a square $x$ is targeted is invariant of the order in which the earthquakes occur. 
\end{theorem}
\begin{proof}
Given the original configuration of the board, let $A_i$ be the number of chips on square $i$. We know that square $i$ will be hit by at least 
\begin{equation*}
\lfloor \frac{A_i}{4} \rfloor
\end{equation*}
earthquakes in order to stabilize. Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$. We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a square that starts with $A_i$ chips on it. Thus using Lemma ~\ref{shiftlemma}, we can move these earthquakes to the front of the sequence without changing the final positioning of the board. We do this for all squares $i$.

The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the squares.

Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the squares. We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes. We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).

Thus we have transformed our starting sequence into canonical sequence that is completely determined by the state of the initial board. Therefore we can do this for any other sequence that stabilizes the board and get the same sequence. This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them). 
\end{proof}

@@

\section{Time until Stability}
\label{Time until Stability}

Another question of interest is how many global earthquakes it would take for a finite board to reach a stable state. We will first focus on the case of the board with only one stack containing $n$ chips.

Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
\begin{definition}
We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes. We also define $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize. 
\end{definition}
Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$, since that many earthquakes will reduce the number of chips on $(0,0)$ to less than $4$.
